Easy as Pi, Volume II: On the Goodput and Energy of Network Coding Schemes in the Raspberry Pi

### 2016-08-16 reviewer_reply.txt

- The presentation of the paper is not very good. The authors did not care very much about the
structure, the language (i.e., page 2 line 42 "Furthermore, in [7], the Kodo library was
introduce." Should be introduced....)

We have corrected all the mentioned and other found typos by proof-reading the manuscript.

- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
Reviewer 3:
- - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

- The setting is limited to a network without intermediate nodes and without packet loss
We considered this scenario to measure the code performance on device basis. For example, adding
the effect of packet losses would not permit to estimate the energy consumption on a given device
since there is energy wasted on possibly lost packets in a given setting.

- Usually RLNC is not a practical scheme due to its high storage cost at the intermediate
nodes when the number of packets is large.
For RLNC the amount of collected packets needed to decode would be similar to any other erasure
code the only difference is that they are coded and some overhead due to the coding coefficients
must be hold. Still, in the study we open the possibility to reduce the complexity by using sparse
codes.

- Regarding typos:
We have corrected all the mentioned and other found typos as suggested by proof-reading the
manuscript and changed the figures in section 4.4 to show the goodput instead of the throughput by
properly labelling the axis and verifying that the shown data is correct.

### 2016-09-29 introduction.tex
Uncommited changes?

% RAW INTRODUCTION: NEED OF DISTRIBUTED NETWORKS AND THE RASPBERRY PI
Due to the advent of the \ac{IoT}, approximately 50 billion devices
ranging from sensors to phones are expected to be connected through
data networks in relatively short period of time \cite{cisco2011forecast}.
This massive deployment requires the design and testing of new
distributed systems that permit to manage the amount of traffic
from the proposed services provided by these devices. Therefore,
development platforms that help to quickly deploy, analyze and
evaluate this type of scenarios are highly desirable for research.
With the emergence of the \ac{Raspi}, a lrelatively powerful low-cost
computer with the size of a credit card, these evaluations are becoming
possible now. This platform has been used as a general purpose hardware
for \ac{IoT} applications as reported in surveys such as
\cite{alfuqaha2015iot}. In these applications, the
\ac{Raspi} might be the sensoring or computing entity (or even both) for
a required task. To achieve this, it can be extended from being a
simple computer using self-designed or already available extension modules.

A benefit of using the \ac{Raspi} as development platform is its large
community of supporters. By running standard operating systems such as
Linux or Windows, this permits to utilize standard, well tested and
reliable tools to administrate and maintain these networks in
a flexible, stable and supported manner which is a major requirement
to make a scalable deployment. Moreover, by enabling system designers to
configure and deploy several devices at the same time, possible deployments
of tens, hundreds or even thousands of Raspberry Pi's would allow to
analyze representative data patterns of the \ac{IoT}. Different use cases of
\ac{IoT} applications employing the \ac{Raspi} as a building block can be found
in the literature. A basic study of the \ac{Raspi} as a possible device
for sensor applications can be found in \cite{vujovic2014raspberry}.
Authors in \cite{kruger2015rapid} consider using the \ac{Raspi} as an
\ac{6LoWPAN} gateway for a set of sensor and mobile devices to an
\ac{IP} version 6 network. In \cite{mahmoud2016sensorian}, a general
purpose sensoring platform for \ac{IoT} is presented using the \ac{Raspi}
as its basic building block. Various interesting \ac{IoT} use cases
are the works presented in \cite{wirz2015enabling,alletto2016indoor,
jalali2016fog,ueyama2014exploiting}. In \cite{wirz2015enabling}, many
\ac{Raspi}s are used as controllable smart devices
with network connectivity that may ubiquously interact with different users,
each represented by a mobile device through a smartphone application.
The study in \cite{alletto2016indoor} considers \ac{Raspi}s as a data
processing unit for disseminating artwork content in smart museums.
An \ac{IoT} setting where the \ac{Raspi} is employed as a nano-server
in distributed storage and computing can be found in \cite{jalali2016fog}.
Finally, in \cite{ueyama2014exploiting}, the authors present
the \ac{Raspi} as the processing entity of an unmanned aereal
vehicle application to increase the resilience of a wireless sensor networks.
However, even though all these advances in the \ac{IoT} area regarding
the exploitation of the \ac{Raspi} capabilities, the application data
is forwarded using former conventional routing methods which may
not satisfy the need of a distributed network for \ac{IoT}
applications as mentioned earlier

@@

Compared to other broadly used coding schemes, such as \ac{LDPC}
codes~\cite{gallager1962low} or Reed-Solomon
codes~\cite{reed1960polynomial}, network coding is a technology that
has been studied and implemented in real systems since the early years of
its conception. A decentralized network code that has been proven
to achieve the multicast capacity with very high probability is
\ac{RLNC} \cite{ho2006random}. Later, a work observing the
benefits of employing \ac{RLNC} in meshed networks is the MORE protocol
addressed in \cite{chachulski2007more}. Shortly afterwards, authors
in~\cite{katti2008xors} showed the performance of an implementation
of the COPE protocol for the two-way relay channel
in a wireless network, which relied on minimalistic coding and obtaining
gains over a forwarding scheme. Later, the work
in~\cite{pedersen2008implementation} used commercially available
Symbian OS mobile phones to implement network coding in a
\ac{D2D} cooperation-based application. Furthermore,
in~\cite{kodo2011pedersen}, the Kodo library was introduced. Kodo is
a C++11 network coding library intended to make network coding
basic functionalities easily available for both the research community
and commercial entities. Based on Kodo, the CATWOMAN
protocol~\cite{hundeboll2012catwoman} is implemented on top of the
BATMAN protocol~\cite{johnson2008simple} for WiFi multi-hop meshed
networks. It uses some of the intuition from COPE, but it is deployed
within the X topology with overhearing links. Its source code is available
as open source in the Linux kernel. Moreover, many other successful
implementations have been tested on real world systems such as
found in~\cite{pahlevani2013playncool,krigslund2013core,
paramanathan2013leanandmean,seferoglu2011i2nc}. For the Raspberry Pi device, an
evaluation of \ac{RLNC} can be found in \cite{paramanathan2014sharing}.
However, this evaluation focused particularly on observing the achievable
speeds only for \ac{RLNC} with different configurations of
the code parameters. In this previous work, a relevant practical aspect that
was not evaluated was the use of hardware acceleration through \ac{SIMD} or
the multi-core capabilities of more advanced \ac{Raspi} models. These features are
becoming more frequent in new processors to largely increase their computational
power for the upcoming demand. In this sense, the Raspberry Pi posses an \ac{ARM}
architecture that could be multi-core as mentioned and also exploits the \ac{SIMD}
feature with the optimized NEON instruction set, but still to the best of our knowledge
there has been no documentation in the literature about these capabilities.

Therefore, in this work we provide detailed measurements of the goodput (processing
speed) and energy consumption of \ac{Raspi} models 1 and 2, when
performing network coding operations with different codecs based on \ac{RLNC}
such as: full dense \ac{RLNC}, multi-core enabled \ac{RLNC}, sparse \ac{RLNC}
and tunable sparse \ac{RLNC}. For these coding schemes, the encoder and decoder
implementations from Kodo are able to detect and make use of the \ac{SIMD}
through the NEON instruction set of the Raspberry Pi by recognizing the \ac{ARM}
architecture with its multicore capabilities. We assess
the \ac{Raspi} performance excluding the effect of packet losses or delays in
order to have a description of the processing and energy consumption for only the
codes in terms of their parameters. To achieve this, we perform a measurement
campaign with the indicated coding schemes and their parameters in various models
of a Raspberry Pi device. Our measurements permit us to characterize the mentioned
metrics of these devices showing that processing peeds of 800 Mbps and processing
energy per bit values of 0.1 nJ are possible. Our work is organized as follows.
Section~\ref{sec:schemes} defines the
coding schemes employed in our study. Later, in Section~\ref{sec:metrics}
we describe the considered metrics and methodology for performance comparison
of the codes deployed in the \ac{Raspi}. In Section~\ref{sec:measurements}, we
show the measurements in the \ac{Raspi} models of the mentioned metrics
providing full discussions about the observed operational regimes and effects.
Final conclusions and future work are reviewed in
Section~\ref{sec:conclusions}.

>>>>

% RAW INTRODUCTION: NEED OF DISTRIBUTED NETWORKS AND THE RASPBERRY PI
Due to the advent of the \ac{IoT}, approximately 50 billion devices
ranging from sensors to phones are expected to be connected through
data networks in relatively short period of time \cite{cisco2011forecast}.
This massive deployment requires the design and testing of new
distributed systems that permit to manage the amount of traffic
from the proposed services provided by these devices. Therefore,
development platforms that help to quickly deploy, analyze and
evaluate this type of scenarios are highly desirable for research.
With the emergence of the \ac{Raspi}, a relatively powerful low-cost
computer with the size of a credit card, these evaluations are becoming
possible now. This platform has been used as a general purpose hardware
for \ac{IoT} applications as reported in surveys such as
\cite{alfuqaha2015iot}. In these applications, the
\ac{Raspi} might be the sensoring or computing entity (or even both) for
a required task. To achieve this, it can be extended from being a
simple computer using self-designed or already available extension modules.

A benefit of using the \ac{Raspi} as development platform is its large
community of supporters. By running standard operating systems such as
Linux or Windows, this permits to utilize standard, well tested and
reliable tools to administrate and maintain these networks in
a flexible, stable and supported manner which is a major requirement
to make a scalable deployment. Moreover, by enabling system designers to
configure and deploy several devices at the same time, possible deployments
of tens, hundreds or even thousands of Raspberry Pi's would allow to
analyze representative data patterns of the \ac{IoT}. Different use cases of
\ac{IoT} applications employing the \ac{Raspi} as a building block can be found
in the literature. A basic study of the \ac{Raspi} as a possible device
for sensor applications can be found in \cite{vujovic2014raspberry}.
Authors in \cite{kruger2015rapid} consider using the \ac{Raspi} as an
\ac{6LoWPAN} gateway for a set of sensor and mobile devices to an
\ac{IP} version 6 network. In \cite{mahmoud2016sensorian}, a general
purpose sensoring platform for \ac{IoT} is presented using the \ac{Raspi}
as its basic building block. Various interesting \ac{IoT} use cases
are the works presented in \cite{wirz2015enabling,alletto2016indoor,
jalali2016fog,ueyama2014exploiting}. In \cite{wirz2015enabling}, many
\ac{Raspi}s are used as controllable smart devices
with network connectivity that may ubiquously interact with different users,
each represented by a mobile device through a smartphone application.
The study in \cite{alletto2016indoor} considers \ac{Raspi}s as a data
processing unit for disseminating artwork content in smart museums.
An \ac{IoT} setting where the \ac{Raspi} is employed as a nano-server
in distributed storage and computing can be found in \cite{jalali2016fog}.
Finally, in \cite{ueyama2014exploiting}, the authors present
the \ac{Raspi} as the processing entity of an unmanned aereal
vehicle application to increase the resilience of a wireless sensor networks.
However, even though all these advances in the \ac{IoT} area regarding
the exploitation of the \ac{Raspi} capabilities, the application data
is forwarded using former conventional routing methods which may
not satisfy the need of a distributed network for \ac{IoT}
applications as mentioned earlier.

@@

Compared to other broadly used coding schemes, such as \ac{LDPC}
codes~\cite{gallager1962low} or Reed-Solomon
codes~\cite{reed1960polynomial}, network coding is a technology that
has been studied and implemented in real systems since the early years of
its conception. A decentralized network code that has been proven
to achieve the multicast capacity with very high probability is
\ac{RLNC} \cite{ho2006random}. Later, a work observing the
benefits of employing \ac{RLNC} in meshed networks is the MORE protocol
addressed in \cite{chachulski2007more}. Shortly afterwards, authors
in~\cite{katti2008xors} showed the performance of an implementation
of the COPE protocol for the two-way relay channel
in a wireless network, which relied on minimalistic coding and obtaining
gains over a forwarding scheme. Later, the work
in~\cite{pedersen2008implementation} used commercially available
Symbian OS mobile phones to implement network coding in a
\ac{D2D} cooperation-based application. Furthermore,
in~\cite{kodo2011pedersen}, the Kodo library was introduced. Kodo is
a C++11 network coding library intended to make network coding
basic functionalities easily available for both the research community
and commercial entities. Based on Kodo, the CATWOMAN
protocol~\cite{hundeboll2012catwoman} is implemented on top of the
BATMAN protocol~\cite{johnson2008simple} for WiFi multi-hop meshed
networks. It uses some of the intuition from COPE, but it is deployed
within the X topology with overhearing links. Its source code is available
as open source in the Linux kernel. Moreover, many other successful
implementations have been tested on real world systems such as
found in~\cite{pahlevani2013playncool,krigslund2013core,
paramanathan2013leanandmean,seferoglu2011i2nc}. For the Raspberry Pi device, an
evaluation of \ac{RLNC} can be found in \cite{paramanathan2014sharing}.
However, this evaluation focused particularly on observing the achievable
speeds only for \ac{RLNC} with different configurations of
the code parameters. In this previous work, a relevant practical aspect that
was not evaluated was the use of hardware acceleration through \ac{SIMD} or
the multicore capabilities of more advanced \ac{Raspi} models. These features are
becoming more frequent in new processors to largely increase their computational
power for the upcoming demand. In this sense, the Raspberry Pi posses an \ac{ARM}
architecture that could be multicore as mentioned and also exploits the \ac{SIMD}
feature with the optimized NEON instruction set, but still to the best of our knowledge
there has been no documentation in the literature about these capabilities.

Therefore, in this work we provide detailed measurements of the goodput (processing
speed) and energy consumption of \ac{Raspi} models 1 and 2, when
performing network coding operations with different codecs based on \ac{RLNC}
such as: full dense \ac{RLNC}, multicore enabled \ac{RLNC}, sparse \ac{RLNC}
and tunable sparse \ac{RLNC}. For these coding schemes, the encoder and decoder
implementations from Kodo are able to detect and make use of the \ac{SIMD}
through the NEON instruction set of the Raspberry Pi by recognizing the \ac{ARM}
architecture with its multicore capabilities. We assess
the \ac{Raspi} performance excluding the effect of packet losses or delays in
order to have a description of the processing and energy consumption for only the
codes in terms of their parameters. To achieve this, we perform a measurement
campaign with the indicated coding schemes and their parameters in various models
of a Raspberry Pi device. Our measurements permit us to characterize the mentioned
metrics of these devices showing that processing peeds of 800 Mbps and processing
energy per bit values of 0.1 nJ are possible. Our work is organized as follows.
Section~\ref{sec:schemes} defines the
coding schemes employed in our study. Later, in Section~\ref{sec:metrics}
we describe the considered metrics and methodology for performance comparison
of the codes deployed in the \ac{Raspi}. In Section~\ref{sec:measurements}, we
show the measurements in the \ac{Raspi} models of the mentioned metrics
providing full discussions about the observed operational regimes and effects.
Final conclusions and future work are reviewed in
Section~\ref{sec:conclusions}.

### 2016-09-29 schemes.tex
Uncommited changes?

A decoder starts to calculate and substract contributions from each
of the pivot elements, e.g. leftmost elements in the main diagonal of
\eqref{eq:coding_matrix}, from top to bottom. The purpose is to obtain
the equivalence  $\textbf{V}$ in its reduced echelon form. The steps
for reducing the matrix by elementary row operations are carried out each time
a linearly independent packet is received. Once in reduced echelon form, packets
can be retrieved by doing a substitution starting from the latest
coded packet. In this way, the amount of elementary operations at the end
of the decoding process is diminished.

@@

In \eqref{eq:recoded_packet}, $w_{ij}$ is the coding coefficient used
in the $i$-th recoded packet and assigned to multiply a previously
coded packet $C_j$. These coding coefficients are again uniformly and
randomly chosen from $GF(q)$. However, these $w_{ij}$'s are not
appended to the previous coding coefficients.  Instead, the system
will update the previous coefficients. Due to the linearity of the
operation, this update reduces to recombining the coding coefficients
equivalent to each original packet with the weight of the
$w_{i,j}$.Therefore, we define the local coding matrix $\textbf{W}$ in
the same way as it was made for $\textbf{V}$. Thus, the local coding
matrix can be written as:
%
\begin{align} \label{eq:local_coding_matrix}
\textbf{W} =
\left[
\begin{array}{c}
        w_1    \\ \hline
        \vdots \\ \hline
        w_g    \\
\end{array}
\right]
=
\left[
\begin{array}{ccc}
        w_{11} & \ldots & w_{1g} \\
        \vdots  & \ddots & \vdots  \\
        w_{g1} & \ldots & w_{gg} \\
\end{array}
\right].
\end{align}
%
With the definitions from \eqref{eq:coding_matrix} and
\eqref{eq:local_coding_matrix}, the recoded packets
$\textbf{R} = \left[R_1 \ldots R_g \right]^T$ are written as
$\textbf{R} = (\textbf{W} \textbf{V}) \textbf{P}$. Here, we recognize
the relationship between original and recoded packets. The resulting
coding matrix is the multiplication of matrices $\textbf{W}$ and
$\textbf{V}$. Denoting ${(\textbf{W} \textbf{V})}_{ij}$ as the element
in the $i$-th row and $j$-th column of $(\textbf{W} \textbf{V})$, this
term is the resulting coding coefficient used to create $R_i$ after
encoding the original packet $P_j$ recoding it locally in an intermediate
node. Finally, the appended coefficients for $R_i$ are
${(\textbf{W} \textbf{V})}_{ik}$ with $k \in [1,2,\ldots,g]$. By doing some
algebra, each ${(\textbf{W} \textbf{V})}_{ij}$ term can be verified to be
computed as:
%
\begin{align}
\label{eq:appended_coded_coefficients}
{(\textbf{W} \textbf{V})}_{ij} = \sum_{k=1}^{g} w_{ik} v_{kj}\ ,\ \forall i,j \in [1,2,\ldots,g] \times [1,2,\ldots,g].
\end{align}
%
This update procedure on the coding coefficients is carried by all the
recoders in a given network therefore allowing any decoder to compute
the original data after Gaussian elimination, regardless of the amount
of times recoding was performed and without incurring in any
additional overhead cost for signaling. Similar to the encoding
operation, any decoder that collects a set of $g$ linearly independent recoded
packets with their respective coefficients, will be able to decode the
data as mentioned before in Section \ref{sssec:decoding}.

@@

The drawback of this scheme is that coded packets from the encoder
become more linearly dependent on each other as the density is reduced.  This
leads to transmission overhead since another coded packet is required
to be sent for every reception of a redundant packet.  Also, this
method may still generate a coded packet, which does not contain any
information. For example, we might find the case where
$\left[\mathbb{B}_1, \ldots, \mathbb{B}_g \right] = \mathbf{0}$ may
happen frequently for low densities. In that case, the encoder
discards the coded packet and tries to generate a new one to avoid the
negative impact on overall system performance.

@@

Therefore, in \ac{TSNC}, a simple procedure for controlling this probability
is to gradually increase the coding density as the \ac{dof}\footnote{We refer
as degrees of freedom to the dimension of the linear span from the coded
packets that a decoder has at a given point} increases, e.g., as the
cardinality of the set of linearly independent coded packets increases. This enables
\ac{TSNC} to significantly reduce the complexity, particularly in the
beginning of the transmission process and also to control the decoding
delay. For \ac{TSNC}, we define the budget, $b \geq g$, to be a target
number of coded packets a transmitter wants to send to a receiver
for decoding a generation of $g$ packets. In some scenarios, we may set
$e$ defined as the excess of linearly dependent packets sent from the encoder.
This helps to also define the budget as $b = g + e$.

>>>>

A decoder starts to calculate and substract contributions from each
of the pivot elements, e.g. leftmost elements in the main diagonal of
\eqref{eq:coding_matrix}, from top to bottom. The purpose is to obtain
the equivalence of $\textbf{V}$ in its reduced echelon form. The steps
for reducing the matrix by elementary row operations are carried out each time
a linearly independent packet is received. Once in reduced echelon form, packets
can be retrieved by doing a substitution starting from the latest
coded packet. In this way, the amount of elementary operations at the end
of the decoding process is diminished.

@@

In \eqref{eq:recoded_packet}, $w_{ij}$ is the coding coefficient used
in the $i$-th recoded packet and assigned to multiply a previously
coded packet $C_j$. These coding coefficients are again uniformly and
randomly chosen from $GF(q)$. However, these $w_{ij}$'s are not
appended to the previous coding coefficients.  Instead, the system
will update the previous coefficients. Due to the linearity of the
operation, this update reduces to recombining the coding coefficients
equivalent to each original packet with the weight of the
$w_{ij}$.Therefore, we define the local coding matrix $\textbf{W}$ in
the same way as it was made for $\textbf{V}$. Thus, the local coding
matrix can be written as:
%
\begin{align} \label{eq:local_coding_matrix}
\textbf{W} =
\left[
\begin{array}{c}
        w_1    \\ \hline
        \vdots \\ \hline
        w_g    \\
\end{array}
\right]
=
\left[
\begin{array}{ccc}
        w_{11} & \ldots & w_{1g} \\
        \vdots  & \ddots & \vdots  \\
        w_{g1} & \ldots & w_{gg} \\
\end{array}
\right].
\end{align}
%
With the definitions from \eqref{eq:coding_matrix} and
\eqref{eq:local_coding_matrix}, the recoded packets
$\textbf{R} = \left[R_1 \ldots R_g \right]^T$ are written as
$\textbf{R} = (\textbf{W} \textbf{V}) \textbf{P}$. Here, we recognize
the relationship between original and recoded packets. The resulting
coding matrix is the multiplication of matrices $\textbf{W}$ and
$\textbf{V}$. Denoting ${(\textbf{W} \textbf{V})}_{ij}$ as the element
in the $i$-th row and $j$-th column of $(\textbf{W} \textbf{V})$, this
term is the resulting coding coefficient used to create $R_i$ after
encoding the original packet $P_j$ recoding it locally in an intermediate
node. Finally, the appended coefficients for $R_i$ are
${(\textbf{W} \textbf{V})}_{ik}$ with $k \in [1,2,\ldots,g]$. By doing some
algebra, each ${(\textbf{W} \textbf{V})}_{ij}$ term can be verified to be
computed as:
%
\begin{align}
\label{eq:appended_coded_coefficients}
{(\textbf{W} \textbf{V})}_{ij} = \sum_{k=1}^{g} w_{ik} v_{kj}\ ,\ \forall i,j \in [1,2,\ldots,g] \times [1,2,\ldots,g].
\end{align}
%
This update procedure on the coding coefficients is carried by all the
recoders in a given network therefore allowing any decoder to compute
the original data after Gaussian elimination, regardless of the amount
of times recoding was performed and without incurring in any
additional overhead cost for signaling. Similar to the encoding
operation, any decoder that collects a set of $g$ linearly independent recoded
packets with their respective coefficients, will be able to decode the
data as mentioned before in Section \ref{sssec:decoding}.

@@

The drawback of this scheme is that coded packets from the encoder
become more linearly dependent on each other as the density is reduced.  This
leads to transmission overhead since another coded packet is required
to be sent for every reception of a redundant packet.  Also, this
method may still generate a coded packet, which does not contain any
information. For example, we might find the case where
$\left[\mathbb{B}_1, \ldots, \mathbb{B}_g \right] = \mathbf{0}$ to occur
frequently for low densities. In that case, the encoder
discards the coded packet and tries to generate a new one to avoid the
negative impact on overall system performance.

@@

Therefore, in \ac{TSNC}, a simple procedure for controlling this probability
is to gradually increase the coding density as the \ac{dof} \footnote{We refer
as degrees of freedom to the dimension of the linear span from the coded
packets that a decoder has at a given point} increases, e.g., as the
cardinality of the set of linearly independent coded packets increases. This enables
\ac{TSNC} to significantly reduce the complexity, particularly in the
beginning of the transmission process and also to control the decoding
delay. For \ac{TSNC}, we define the budget, $b \geq g$, to be a target
number of coded packets a transmitter wants to send to a receiver
for decoding a generation of $g$ packets. In some scenarios, we may set
$e$ defined as the excess of linearly dependent packets sent from the encoder.
This helps to also define the budget as $b = g + e$.
